Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-09-01 paper/main.tex
spivak reworking of typing bundles

\IEEEPARstart{V}isualizations are expected to be a faithful translation of data, meaning that inherent structure in the data should be present in the visualization; therefore the components of visualization software libraries that translate data to graphics are also expected to be structure preserving. This inherent structure can be quite complex, such as when the data is high dimensional, multivariate, deeply nested, or encoded in a distributed manner. We propose that we can rigorously specify what structure the library components are expected to preserve using category theory. We also propose that these specifications can generalize to most structure types by extending previous work on encoding  inherent structure in a consistent manner using mathematical structures from algebraic topology. These algebraic structures can be translated into analogous programmatic types for encoding structure, while the use of category theory facilitates developing a functional design framework that describes function composition in terms of functional algebraic operations. The typing system and composition inherent in a functional design encourage library developers to build complex visualization components from simple verifiable parts that have few side effects \cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These categorically specified components could be built as a standalone library and integrated into existing libraries because they are inherently self-contained. Furthermore we see the application of these ideas in low level visualization library design as our motivation, and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib. The contribution of this paper is a mathematical framework for describing structure preserving visualization components in a manner that is generalizable to different types of inherent structure. This framework also provides guidance for the construction and testing of structure preserving visualization library components.

@@

The topological structure of the data can be expressed programmatically by constructing data types that encapsulate the class of topological structure-.i.e point, line, plane, network, etc. We propose that these data types can be formally specified as objects of a category. 
\begin{definition} 
   An \textbf{category} $\mathcal{C}$ consists of the following \textit{data}: 
\begin{enumerate}
  \item a collection of \textit{objects} $X \in \mathcal{Ob}(\mathcal{C})$
  \item for every pair of objects $X, Y \in \mathcal{Ob}(\mathcal{C})$, a set of \textit{morphisms} $X \xrightarrow{f} Y \in Hom_{\mathcal{C}}(X, Y)$
  \item for every object $X$, a distinct \textit{identity morphism} $X \xrightarrow {id_x} X$ in $Hom_{\mathcal{C}}(X, X)$
  \item a \textit{composition function} $f \in Hom_{\mathcal{C}}(X, Y) \times  g \in Hom_{\mathcal{C}}(Y, Z) \rightarrow g \circ f \in Hom_{\mathcal{C}}(X, Z)$
\end{enumerate}
such that 
\begin{enumerate}
  \item \textit{unitality:} for every morphism $ X \xrightarrow{f} Y$, $f \circ id_x = f = id_y \circ f$
  \item \textit{associativity:} if any three morphisms $f, g, h$ are composable, 
    \begin{equation*}
      \begin{tikzcd}
        X \arrow[r, "f"] \arrow[rrr, "h\circ(g\circ f) = (h\circ g)\circ f"', bend right, dashed] & Y  \arrow[r, "g"] & Z \arrow[r, "h"] & W
        \end{tikzcd}
  \end{equation*}
  then they are associative such that $h\circ(g\circ f) = (h \circ g) \circ f$  \cite{lawvere2009conceptual,riehlCategoryTheoryContext,maclaneCategoriesWorkingMathematician2013,fongInvitationAppliedCategory2019}. 
  \end{enumerate}
\end{definition}

We encapsulate the topological structure of the data as a category $\mathcal{\dbasec}$. The standard construction of a category from a topological space is that it has open set objects $\opensetc$ and inclusion morphisms $\opensetc_i \xrightarrow{\iota} \opensetc_j$ such that $\opensetc_i \subseteq \opensetc_j$\cite{riehlCategoryTheoryContext}. The composability property expresses that inclusion is transitive, while associativity expresses that the inclusion functions can be curried in various equivalent groupings. By formally specifying the properties of the topological structure datatypes as $\mathcal{\dbase}$, we can express that these are the properties that are required as part of the implementation of the data type objects.

\subsubsection{Data Fields: Fiber Space \dfiber}
\label{sec:atct:fb:fiber}
As mentioned in \autoref{sec:related-work:equivariance}, visualization researchers traditionally describe equivariance as the preservation of field structure, which is based on the field type. Spivak shows that data typing can be expressed in a categorical framework in his fiber bundle formulation of tables in relational databases \cite{spivakDatabasesAreCategories2010,spivakSimplicialDatabases2009}. In this work, we adopt Spivak's definitions of \textit{type specification}, \textit{schema}, \textit{record} because that allows us to use a dimension agnostic named typing system for the fields of our dataset that is consistent with the abstraction we are using to express the continuity. Spivak introduces a \textit{type specification} as a bundle map $\pi: \mathscr{U} \rightarrow \textbf{DT}$. The base space $\textbf{DT}$ is the set of data types, and the total space $\mathscr{U}$ is the disjoint union of the domains of those data types. Given an arbitrary data type $T \in \textbf{DT}$, the preimage $ \pi^{-1}(T) \subset \mathscr{U}$ is the domain of $T$ and $x \in \pi^{-1}(T)$ is an object of type $T$\cite{spivakSimplicialDatabases2009}. For example, if $T=\texttt{int}$, then the image $\pi^{-1}(\texttt{int}) = \mathbb{Z} \subset \mathscr{U}$ is the set of all integers. Since many fields can have the same datatype, Spivak formally defines a mapping from field name to field data type, akin to a database schema \cite{ullmanFirstCourseDatabase2008}. According to Spivak, a \textit{schema} consists of a pair $(\fnames, \sigma)$ where $\fnames$ is the set of field names and $\sigma: \fnames \rightarrow \textbf{DT}$ is a function from field name to field data type\cite{spivakSIMPLICIALDATABASESa}. The function $\sigma$ is composed with $\pi$ such that the bundle  $\pi_{\sigma}:\mathscr{U}_{\sigma} \rightarrow \fnames$ associates a field name $c \in C$ with the corresponding domain for that data type $\mathscr{\U}_{\sigma}$. 
\begin{definition} A \textbf{record} is a function $\delement: \fnames \rightarrow \mathscr{U}_{\sigma}$ and the set of records on $\pi_{\sigma}$ is denoted $\Gamma^{\pi}(\sigma)$. Records must return an object of type $\sigma(\fname) \in \textbf{DT}$ for each field $c \in C$.
\end{definition}
Spivak then describes tables as sections $\dsection: \dbase \rightarrow \Gamma^{\pi}(\sigma)$ from a key space $\dbase$ to the set of all possible records $\Gamma^{\pi}(\sigma)$ on the schema bundle, and his notion of a table generalizes to our notion of a data container.

@@

As with the base space category $\mathcal{\dbase}$, we propose a fiber category $\mathcal{\dfiberc}$ to encapsulate the field types of the data. The fiber category has a single object $\dfiberc$ of an arbitrary type and morphisms on the fiber object $\dfunctc \in Hom(\dfiberc, \dfiberc)$. The fiber category $\mathcal{\dfiberc}$ is a monoidal category, meaning that it is a category equipped with a bifunctor $\otimes: \mathcal{\dfiber} \times \mathcal{\dfiber} \rightarrow \mathcal{\dfiber}$. The bifunctor provides a method for combining fibers, thereby allowing us to express fields that contain multityped values. For example, wind can be represented as two fields $\dfiber_{speed} \times \dfiber_{direction}$ or a composite fiber field $\dfiber_{speed} \otimes \dfiber_{direction} = \dfiber_{wind}$. The $\otimes$ encapsulates both the sets associated with each fiber $\mathbb{R} \times \mathbb{R} = \mathbb{R}^{2}$ and the morphisms associated with each functor $(\dfunct_{speed}, \dfunct_{direction}) = \dfunct_{wind}$. Defining the field schema as a monoidal category allows us to formally express the structure of the field, as defined by its sets and functions, at different levels of composition as needed by different visualization tasks.

\subsubsection{Data: Section} 
\label{sec:atct:fb:sections}
We encode data as a \textcolor{section}{section} \dsectionc\ of a bundle because this allows us to incorporate the topology and field types in the data definition. We can define these section functions locally, meaning over a specific open subset \openset\ of \dbase\ 
\begin{equation}
  \label{eq:atct:fb:sections}
  \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}} \coloneqq \big\{\dsectionc: \opensetc\rightarrow \dtotalc\restriction_{\opensetc} \; \bigm{\vert} \pi(\dsectionc(\dbasepointc)) = \dbasepointc\;for\, all\; \dbasepointc \in \opensetc \big\} 
\end{equation}
such that each section function $\dsection: \dbasepoint \mapsto \delement$ maps from a point $\dbasepoint \in \openset$ to a record in the fiber space $\delement \in \dfiber_{\dbasepoint}$ over that point. Bundles can have multiple sections, as denoted by $\Gamma(\openset, \dtotal\restriction{\openset})$. We can therefore model data as structures that map from an index like point $\dbasepoint$ to a data record $\delement$, and encapsulate multiple datasets with the same fiber and base space as different sections of the same bundle.

In a trivial bundle all fibers are equal $\dfiberc \equiv \dfiberc_{\dbasepointc}\;\forall \dbasepointc \in \dbase$, which means that $\dtotalc = \dbasec \times \dfiberc$ . This allows us to define global sections $\dsection: \dbase \rightarrow \dfiber \in \Gamma(\dbase, \dfiber)$ which we translate into a data signature of the form  
\begin{equation}
  \textcolor{section}{\texttt{dataset}}: \textcolor{base}{\texttt{topology}} \rightarrow \textcolor{fiber}{\texttt{field}}
\end{equation}
where $\dsection=\texttt{dataset}$, $\dbase=\texttt{topology}$ and $\dfiber=\texttt{fields}$. This type signature provides a method of explicitly stating the topology and field type of the data and generalizes to almost any topology and fiber type, provided that the total space is trivial.

When the total space is non-trivial, we can use the fiber bundle property of local-triviality to generate type signatures. In a non-trivial bundle, fibers are isomorphic $\dfiber \simeq \dfiber_{\dbasepoint}\;\forall \dbasepoint \in \dbase$, which means there is no single fiber type $\dfiber$. Instead, we can define sections locally $\dsection\restriction_{\openset} \in \Gamma(\openset_{\dbasepoint}, \dtotal\restriction_{\openset_{\dbasepoint}})$, which means a section defined only on an open neighborhood\footnote{An open set that surrounds a point \dbasepoint\ is called an open neighborhood $\openset_{\dbasepoint}$.} $\dbasepoint \in \openset \in \dbase$. This is because all fiber bundles are locally trivializable, $\dtotal\restriction_{\openset_{k}} = \openset_{k}\times \dfiber$ such that there's always a subset where the fibers are equal. This means almost all data can be expressed as a collection of local sections $\{\dsection\restriction_{\openset_{\dbasepoint}}| \dbasepoint \in \dbase \}$ which translate to a set of signatures 
\begin{equation}
  \begin{split}
\{\textcolor{section}{\texttt{data-subset}}: \textcolor{base}{\texttt{topology}} \rightarrow \textcolor{fiber}{\texttt{fields}} \\
 s.\;t.\;\textcolor{section}{\texttt{data-subset}} \in \texttt{dataset}\}  
  \end{split}
\end{equation}
that can be glued together on overlaps to generate the dataset. This gluing can be implemented as part of the data container, which we discuss in \autoref{sec:atct:sheaves}.

\subsubsection{Examples}
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{tbundle.png}
  \caption{The mobius band is a non-trivial total space because the twist means that the fibers are sometimes [-1, 1] and other times [1,-1]. The mobius band can be represented as two trivial bundles, one with fiber [-1,1] and one with fiber [1,-1] that can be glued on the edges into the band. The gluing rules ensure that the arrows on the edges go in the same direction.\label{fig:atct:nontrivialbundle}}
\end{figure}

@@

\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{bibliography}

% biography section 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

>>>>

\IEEEPARstart{V}isualizations are expected to be a faithful translation of data, meaning that inherent structure in the data should be present in the visualization; therefore the components of visualization software libraries that translate data to graphics are also expected to be structure preserving. This inherent structure can be quite complex, such as when the data is high dimensional, multivariate, deeply nested, or encoded in a distributed manner. We propose that we can rigorously specify what structure the library components are expected to preserve using category theory. We also propose that these specifications can generalize to most structure types by extending previous work on encoding  inherent structure in a consistent manner using mathematical structures from algebraic topology. These algebraic structures can be translated into analogous programmatic types for encoding structure, while the use of category theory facilitates developing a functional design framework that describes function composition in terms of functional algebraic operations. The typing system and composition inherent in a functional design encourage library developers to build complex visualization components from simple verifiable parts that have few side effects\cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These categorically specified components could be built as a standalone library and integrated into existing libraries because they are inherently self-contained. Furthermore we see the application of these ideas in low level visualization library design as our motivation, and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib. The contribution of this paper is a mathematical framework for describing structure preserving visualization components in a manner that is generalizable to different types of inherent structure. This framework also provides guidance for the construction and testing of structure preserving visualization library components.

@@

The topological structure of the data can be expressed programmatically by constructing data types that encapsulate the class of topological structure-.i.e point, line, plane, network, etc. We propose that these data types can be formally specified as objects of a category. 
\begin{definition} 
   An \textbf{category} $\mathcal{C}$ consists of the following \textit{data}: 
\begin{enumerate}
  \item a collection of \textit{objects} $X \in \mathcal{Ob}(\mathcal{C})$
  \item for every pair of objects $X, Y \in \mathcal{Ob}(\mathcal{C})$, a set of \textit{morphisms} $X \xrightarrow{f} Y \in Hom_{\mathcal{C}}(X, Y)$
  \item for every object $X$, a distinct \textit{identity morphism} $X \xrightarrow {id_x} X$ in $Hom_{\mathcal{C}}(X, X)$
  \item a \textit{composition function} $f \in Hom_{\mathcal{C}}(X, Y) \times  g \in Hom_{\mathcal{C}}(Y, Z) \rightarrow g \circ f \in Hom_{\mathcal{C}}(X, Z)$
\end{enumerate}
such that 
\begin{enumerate}
  \item \textit{unitality:} for every morphism $ X \xrightarrow{f} Y$, $f \circ id_x = f = id_y \circ f$
  \item \textit{associativity:} if any three morphisms $f, g, h$ are composable, 
    \begin{equation*}
      \begin{tikzcd}
        X \arrow[r, "f"] \arrow[rrr, "h\circ(g\circ f) = (h\circ g)\circ f"', bend right, dashed] & Y  \arrow[r, "g"] & Z \arrow[r, "h"] & W
        \end{tikzcd}
  \end{equation*}
  then they are associative such that $h\circ(g\circ f) = (h \circ g) \circ f$  \cite{lawvere2009conceptual,riehlCategoryTheoryContext,maclaneCategoriesWorkingMathematician2013,fongInvitationAppliedCategory2019}. 
  \end{enumerate}
\end{definition}
We encapsulate the topological structure of the data as a category $\mathcal{\dbasec}$. The standard construction of a category from a topological space is that it has open set objects $\opensetc$ and inclusion morphisms $\opensetc_i \xrightarrow{\iota} \opensetc_j$ such that $\opensetc_i \subseteq \opensetc_j$\cite{riehlCategoryTheoryContext}. The composability property expresses that inclusion is transitive, while associativity expresses that the inclusion functions can be curried in various equivalent groupings. By formally specifying the properties of the topological structure datatypes as $\mathcal{\dbase}$, we can express that these are the properties that are required as part of the implementation of the data type objects.

\subsubsection{Data Fields: Fiber Space \dfiber}
\label{sec:atct:fb:fiber}
As mentioned in \autoref{sec:related-work:equivariance}, visualization researchers traditionally describe equivariance as the preservation of field structure, which is based on the field type. Spivak shows that data typing can be expressed in a categorical framework in his fiber bundle formulation of tables in relational databases \cite{spivakDatabasesAreCategories2010,spivakSimplicialDatabases2009}. In this work, we adopt Spivak's definitions of \textit{type specification}, \textit{schema}, and \textit{record} because that allows us to use a dimension agnostic named typing system for the fields of our dataset that is consistent with the abstraction we are using to express the continuity. Spivak introduces a \textit{type specification} as a bundle map $\pi: \mathscr{U} \rightarrow \textbf{DT}$. The base space $\textbf{DT}$ is a set of data types $T \in \textbf{DT}$ and the total space $\mathscr{U}$ is the disjoint union of the domains of each type \[\mathscr{U} = \bigsqcup_{T \in \textbf{DT}} \pi^{-1}(T)\] such that each element $x$ in the domain $\pi^{-1}(T)$ is one possible value of an object of type $T$ \cite{spivakSimplicialDatabases2009}. For example, if $T=\texttt{int}$, then the image $\pi^{-1}(\texttt{int}) = \mathbb{Z} \subset \mathscr{U}$ is the set of all integers and $x=3 \in \mathbb{Z}$ is the value of one $\texttt{int}$ object.

Since many fields can have the same datatype, Spivak formally defines a mapping from field name to field data type, akin to a database schema \cite{ullmanFirstCourseDatabase2008}. According to Spivak, a \textit{schema} consists of a pair $(\fnames, \sigma)$ where $\fnames$ is the set of field names and $\sigma: \fnames \rightarrow \textbf{DT}$ is a function from field name to field data type\cite{spivakSimplicialDatabases2009}. The function $\sigma$ is composed with $\pi$ such that $\pi^{-1}(\sigma(C)) \subseteq \mathscr{U}$; this composition induces a domain bundle $\pi_{\sigma}:\mathscr{U}_{\sigma} \rightarrow \fnames$ that associates a field name $c \in C$ with its corresponding domain $\pi^{-1}_{\sigma}(C) \subseteq \mathscr{U}_{\sigma}$. 
\begin{definition} A \textbf{record} is a function $\delement: \fnames \rightarrow \mathscr{U}_{\sigma}$ and the set of records on $\pi_{\sigma}$ is denoted $\Gamma^{\pi}(\sigma)$. Records must return an object of type $\sigma(\fname) \in \textbf{DT}$ for each field $c \in C$.
\end{definition}
Spivak then describes tables as sections $\dsection: \dbase \rightarrow \Gamma^{\pi}(\sigma)$ from a key space $\dbase$ to the set of all possible records $\Gamma^{\pi}(\sigma)$ on the schema bundle, and his notion of a table generalizes to our notion of a data container.

@@

As with the base space category $\mathcal{\dbase}$, we propose a fiber category $\mathcal{\dfiberc}$ to encapsulate the field types of the data. The fiber category has a single object $\dfiberc$ of an arbitrary type and morphisms on the fiber object $\dfunctc \in Hom(\dfiberc, \dfiberc)$. The fiber category $\mathcal{\dfiberc}$ is a monoidal category, meaning that it is a category equipped with a bifunctor $\otimes: \mathcal{\dfiber} \times \mathcal{\dfiber} \rightarrow \mathcal{\dfiber}$. The bifunctor provides a method for combining fibers, thereby allowing us to express fields that contain multityped values. For example, wind can be represented as two fields $\dfiber_{speed} \times \dfiber_{direction}$ or a composite fiber field $\dfiber_{speed} \otimes \dfiber_{direction} = \dfiber_{wind}$. The $\otimes$ encapsulates both the sets associated with each fiber $\mathbb{R} \times \mathbb{R} = \mathbb{R}^{2}$ and the morphisms associated with each functor $(\dfunct_{speed}, \dfunct_{direction}) = \dfunct_{wind}$. Defining the field schema as a monoidal category allows us to formally express the structure of the field, as defined by its sets and functions, at different levels of composition as needed by different visualization tasks.
 \
\subsubsection{Data: Section} 
\label{sec:atct:fb:sections}
We encode data as a \textcolor{section}{section} \dsectionc\ of a bundle because this allows us to incorporate the topology and field types in the data definition. We can define these section functions locally, meaning that the section is (piece-wise) continuous over a specific open subset \openset\ of \dbase\
\begin{equation}
  \label{eq:atct:fb:sections}
  \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}} \coloneqq \big\{\dsectionc: \opensetc\rightarrow \dtotalc\restriction_{\opensetc} \; \bigm{\vert} \pi(\dsectionc(\dbasepointc)) = \dbasepointc\;for\, all\; \dbasepointc \in \opensetc \big\} 
\end{equation}
such that each section function $\dsection: \dbasepoint \mapsto \delement$ maps from each point $\dbasepoint \in \openset$ to a corresponding record in the fiber space $\delement \in \dfiber_{\dbasepoint}$ over that point. Bundles can have multiple sections, as denoted by $\Gamma(\openset, \dtotal\restriction{\openset})$. We can therefore model data as structures that map from an index like point $\dbasepoint$ to a data record $\delement$, and encapsulate multiple datasets with the same fiber and base space as different sections of the same bundle.

In a trivial bundle, the fibers are equal $\dfiberc =  \dfiberc_{\dbasepointc}$ over all points $\dbasepointc \in \dbase$  in the base space such that the total space is the product of the fiber and base space $\dtotalc = \dbasec \times \dfiberc$. This allows us to define global sections $\dsection: \dbase \rightarrow \dfiber \in \Gamma(\dbase, \dfiber)$ which we translate into a data signature of the form  
\begin{equation}
  \textcolor{section}{\texttt{dataset}}: \textcolor{base}{\texttt{topology}} \rightarrow \textcolor{fiber}{\texttt{field}}
\end{equation}
where $\dsection=\texttt{dataset}$, $\dbase=\texttt{topology}$ and $\dfiber=\texttt{fields}$. This type signature provides a method of explicitly stating the topology and field type of the data and generalizes to almost any topology and fiber type, provided that the total space is trivial.

When the total space is non-trivial, we can use the fiber bundle property of local-triviality to generate type signatures. In a non-trivial bundle fibers are isomorphic $\dfiber \simeq \dfiber_{\dbasepoint}$ for all points $\dbasepoint \in \dbase$, which means that there may be a \textit{twist} in the bundle that prevents defining global sections \cite{hatcherAlgebraicTopology2002}. As mentioned in \autoref{def}

Fiber bundles are locally trivializable, $\dtotal\restriction_{\openset_{k}} = \openset_{k}\times \dfiber$ such that there's always a subset where the fibers are equal.

Instead, we can define sections locally $\dsection\restriction_{\openset} \in \Gamma(\openset_{\dbasepoint}, \dtotal\restriction_{\openset_{\dbasepoint}})$, which means a section defined only on an open neighborhood\footnote{An open set that surrounds a point \dbasepoint\ is called an open neighborhood $\openset_{\dbasepoint}$.} $\dbasepoint \in \openset \in \dbase$. 
 This means almost all data can be expressed as a collection of local sections $\{\dsection\restriction_{\openset_{\dbasepoint}}| \dbasepoint \in \dbase \}$ which translate to a set of signatures 
\begin{equation}
  \begin{split}
\{\textcolor{section}{\texttt{data-subset}}: \textcolor{base}{\texttt{topology}} \rightarrow \textcolor{fiber}{\texttt{fields}} \\
 s.\;t.\;\textcolor{section}{\texttt{data-subset}} \in \texttt{dataset}\}  
  \end{split}
\end{equation}
that can be glued together on overlaps to generate the dataset. This gluing can be implemented as part of the data container, which we discuss in \autoref{sec:atct:sheaves}.

\subsubsection{Examples}
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{tbundle.png}
  \caption{The mobius band is a non-trivial total space because the twist means that the fibers are sometimes [-1, 1] and other times [1,-1]. The mobius band can be represented as two trivial bundles, one with fiber [-1, 1] and one with fiber [1,-1] that can be glued on the edges into the band. The gluing rules ensure that the arrows on the edges go in the same direction.\label{fig:atct:nontrivialbundle}}
\end{figure}

@@

\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{bibliography}
% biography section 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:
